Splitting of the homology of the punctured mapping class group

### 2018-10-29 Actiongg.tex
Finished proving the main theorem!

The construction of $\psi_m$ is rather long and technical and involves a few definitions.
\begin{defn}
 \label{defn:D}
 Let $\D\subset\mrS$ be the open square $(1/4;3/4)\times(1/2,1)$, using the model $\T(\S)$:
$\D$ is an open disc in $\mrS$ near $\partial\S$ and disjoint from all $\U_i$'s and $\V_i$'s.

In the following it is useful to redefine $\gg$ as the group of isotopy classes
of diffeomorphisms of $\S$ that fix both $\partial \S$ and $\D$ pointwise: this is 
not a problem, since the inclusion
\[
\Diff(\S;\D\cup\partial\S)\subset \Diff(\S;\partial\S)
\]
induces a bijection on the set of path-components.
\end{defn}

@@

We have the following diagram of $\gg$-equivariant maps, where the full arrows are those
that we have already constructed, and we still have to prove the existence of the dashed arrows
\begin{equation}
\label{eq:fulldasheddiagram}
\begin{tikzcd}[column sep=8em,row sep=5em]
  \ZcC{m} \ar[r,"\pr_m",two heads]\ar[d,dashed, swap, "\tpsi_m"]\ar[dr,"\j_m",near start]
  & \Sym_m(\H)\ar[d,dashed ]\ar[dl,dashed,very near start,"\psi_m"]\\
  H_m(\cms)\ar[r,swap,"\pa{\iota_m}*"hook] & H_m(\cmsD)
 \end{tikzcd}
\end{equation}

@@

Theorem \ref{thm:Hbms*as*ggrep} is then proved in all bidegrees $(*,m)$ with $*=m$.

\subsection{General bidegrees $(m-l,m)$.} In this subsection we fix a bidegree $(m-l,m)$
with $0\leq l\leq m$.
Our aim is to prove theorem \ref{thm:Hbms*as*ggrep} for this bidegree $(m-l,m)$.
\begin{defn}
%  Let $\tilde\D'$ be the open trapeziod contained in $(0,1)^2\subset\mrS$ which is delimited
%  by the four points $(1/8;1)$, $(1/4;1/2)$, $(3/4;1/2)$ and $(7/8;1)$ (SEE PICTURE); in particular
%  $\D\subset\D'$.
 Let $\S'\subset\S$ be the closure in $\S$ of $\S\setminus\bar\D$, where $\bar\D$ is the closure of $\D$ in $\S$.
 Notice that $\S'$ is also a surface of type $\sg$.

 For all $m$ we define configuration spaces $C_m(\D)$ and $C_m(\S')$ as in definition \ref{defn:cms},
 using $\D$ and $\mrS'$ instead of $\mrS$, respectively.

 For every splitting $m=p+(m-p)$ there is an open embedding
 \[
 \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
 \]
 given by taking the union of configurations:
 \[
  \mu(\set{P_1,\dots,P_p};\set{P'_1,\dots,P_{m-p}})=\set{P_1,\dots,P_p,P'_1,\dots,P'_{m-p}}\in\cms.
 \]
%  For the rest of the section let $\gg$ denote the group of isotopy classes of diffeomorphisms
%  of $\S$ relative to $\partial\S\cup\D'$: in symbols $\gg=\pi_0\pa{\Diff(\S;\partial\S\cup\D'}$.
\end{defn}
For all $0\leq p\leq m$, the group $\Diff(\S;\partial\S\cup\D)$ acts both on $C_p(\D)\times C_{m-p}(\S')$
and on $\cms$, and the map $\mu$ is equivariant with respect to this action;
hence, using the K\"{u}nneth formula, there is an induced $\gg-$equivariant map in homology
\begin{equation}
 \label{eq:mu*}
 \mu_*\colon H_{p-l}\pa{C_p(\D)}\otimes H_{m-p}\pa{C_{m-p}(\S')}\to H_{m-l}\pa{\cms}.
\end{equation}

@@

We will prove the following lemma, from which theorem \ref{thm:Hbms*as*ggrep} easily follows:
\begin{lem}
 \label{lem:oplussplitting}
For all $l\leq p\leq m$ the map $\mu_*$ in equation \ref{eq:mu*} is injective, and the collection
of all these maps yields a splitting
\[
 H_{m-l}(\cms)=\bigoplus_{p=l}^m H_{p-l}\pa{C_p(\D)}\otimes H_{m-p}\pa{C_{m-p}(\S')}
\]
\end{lem}
\begin{proof}
Notice that the statement of the lemma doesn't depend on the the action of $\gg$:
we have a map between two $\Z_2-$vector spaces, we already know that it is $\gg-$equivariant,
we only need to show that it is a bijection.
% 
% We can equivalently consider the maps in cohomology
%  \[
%    \mu^*\colon  H^{m-l}\pa{\cms}\to H^p\pa{C_{m-p}(\S')}\otimes H^{p-l}\pa{C_p(\D)},
%  \]
% and show that they are surjective for $0\leq p\leq m-l$ and together exhibit a product splitting
% \[
%   H^{m-l}(\cms)=\prod_{p=0}^{m-l} H^p\pa{C_{m-p}(\S')}\otimes H^{p-l}\pa{C_p(\D)}.
% \]
% Notice that the natural target for the cohomology map $\mu^*$ is the whole group
% \[
%  H^{m-l}\pa{C_{m-p}(\S')\times C_p(\D)}\simeq \prod_{q=0}^p H^q\pa{C_{m-p}(\S')}\otimes H^{m-l-q}\pa{C_p(\D)},
% \]
% where we have used the K\"unneth formula; we consider here, by abuse of notation,
% the composition of this map with the projection
% on the factor corresponding to $q=p$.
% 
% Using Poincaré-Lefschetz duality and again the K\"unneth formula we reduce to show that the maps
% \[
% \mu^c_* \colon \tH_{m-l}(\cms^c)\to \tH_p\pa{C_{m-p}(\S')^c}\otimes \tH_{p-l}\pa{C_p(\D)^c}
% \]
% are surjective and exhibit $\tH_{m-l}(\cms^c)$ as a product
% \[
% \tH_{m-l}(\cms^c)=\prod_{p=0}^{m-l}\tH_p\pa{C_{m-p}(\S')^c}\otimes \tH_{p-l}\pa{C_p(\D)^c}.
% \]
% Here a few explanations are required:
% we are considering the map $\mu^c\colon \cms^c\to \pa{C_{m-p}(\S')\times C_p(\D)}^c$ which collapses the complement
% of the open subspace $C_{m-p}(\S')\times C_p(\D)\subset\cms$ to the point at infinity; it induces a map
% $\mu^c_*\colon \tH_{m-l}(\cms^c) \to \tH_{m-l}\pa{\pa{C_{m-p}(\S')\times C_p(\D)}^c}$;
% applying Poincaré-Lefschetz duality
% and the K\"unneth formula to the second group we get an isomorphism
% \[
%  \tH_{m-l}\pa{\pa{C_{m-p}(\S')\times C_p(\D)}^c}\simeq \prod_{q=0}^p \tH_q\pa{C_{m-p}(\S')}\otimes\tH_{m-l-q}\pa{C_p(\D)};
% \]
% we are considering the composition of the map $\mu^c_*$ with the projection onto the factor $q=p$, and by
% abuse of notation we call this map simply $\mu^c_*$.
%

Fix $l\leq p\leq m$ and let $[a]=\prod_{j=0}^{\infty}(Q^j\epsilon)^{\alpha_j}$ be a generator of
$H_{p-l}(C_p(\D))$, i.e., $\sum_j\alpha_j=l$ and $\sum_j2^j\alpha_j=p$; let also
$\beta=\prod_{i=1}^g \u_i^{u_i}\v_i^{v_i}$ be a generator of $H_{m-p}(C_{m-p}(\S'))$, using the
isomorphism proved in the previous subsection, i.e., $\sum_i(u_i+v_i)=m-p$. Here $a$ and
$b$ are chosen singular cycles representing the homology classes, with $a$ supported on $\D$
and $b$ supported on $\S'$.

Then $a\otimes b$ represents a generator of $H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S'))$,
and we are interested in the homology class represented by $\mu_*(a\otimes b)$.
We will study
the intersection of $\mu_*([a]\otimes [b])$ with cohomology classes of $\cms$ represented by
generalised symmetric chains in $\cms^c$.

@@

To compute the algebraic intersection
between $\mu_*(a\otimes b)$ and $[p',(\alpha'_j),(u'_i,v'_i)]$ we
consider the map
\[
 \mu^c\colon \cms^c\to \pa{C_p(\D)\times C_{m-p}(\S')}^c
\]
which collapses to $\infty$ the complement in $\cms^c$ of $C_p(\D)\times C_{m-p}(\S')$.

@@

Recall that $\cms^c$ can be filtered according to the norm of cells: a cell $e^{\tup}$ with
$\tup=(l,(x_i),(u_i,v_i))$ has norm $\sum x_i$. In the previous section we just considered
the associated filtration of the reduced chain complex $\tCh_*(\cms^c)$, but now we need
to consider the closed subcomplex $F_p\cms^c\subset\cms^c$ consisting of all cells of norm $\leq p$.

The crucial remark is that $\mu^c$ restricts to a cellular map 
\[
F_p\cms^c\to \pa{C_p(\D)\times C_{m-p}(\S')}^c.
\]
Consider an open cell
cell $e^{\tup}$ for some tuple $\tup=(l, (x_i),(u'_i,v'_i))$ of norm $p'\leq p$:
its intersection with the subspace
$C_p(\D)\times C_{m-p}(\S')$ is empty if $p'<p$, and it is exactly the open product of cells
$e^{\tup'}\times e^{\tup''}$,
where $\tup'=(l,(x_i))$ and $\tup''=(0,(u'_i,v'_i))$, if $p'=p$; therefore
$\mu^c(e^{\tup})$ is $\set{\infty}$ in the first case, and in the second case it
is contained in the union $\set{\infty}\cup e^{\tup'}\times e^{\tup''}$, which is also
contained in the $p'-$skeleton of $\pa{C_p(\D)\times C_{m-p}(\S')}^c$.

We replace the map $\mu^c$ by a cellular approximation that agrees with it on the $p-$skeleton,
and by abuse of notation we still call $\mu^c$ the new map.

Consider now any generalised symmetric chain $|p',(\alpha'_j),(u'_i,v'_i)|$ representing
a class in $H_{m+l}(\cms^c)=H^{m-l}(\cms)$, hence $\sum_j\alpha'_j=l$, and suppose $p'\leq p$.

@@

We consider the collection of all strings of the form $(p,(u_i,v_i)_{i\leq g},(\alpha_j)_j)$
satisfying $\sum\alpha_j=l$ and $\sum \alpha_j2^j=p$; we order these strings lexicographically,
so that in particular $p$ is weakly increasing; we associate
to each string its corresponding class in $H_{m-l}(\cms)$ of the form $\mu_*(a\otimes b)$ and its
corresponding class in $\tH_{m+l}(\cms)$ represented by the corresponding generalised symmetric chain.

Then the matrix of algebraic intersections is an upper-triangular matrix
with $1$'s on the diagonal, and in particular
it is invertible. This shows that the set of classes $\mu_*(a\otimes b)$
is a basis for $H_{m-l}(\cms)$.

@@

One could expect that the basis given by all elements $[a]\otimes [b]\in H_{m-l}(\cms)$
is dual to the basis of classes $[p',(\alpha'_j),(u'_i,v'_i)]\in \tH_{m+l}(\cms^c)$, i.e., the
matrix considered in the end of the previous proof is not only upper-triangular but also
diagonal. This is however not true, as the following example shows...

>>>>

The construction of $\psi_m$ is rather long and technical and involves a few definitions.

@@

We have the following diagram of $\gg$-equivariant maps, where the full arrows are those
that we have already constructed, and we still have to prove the existence of the dashed arrows
\begin{equation}
\label{eq:fulldasheddiagram}
\begin{tikzcd}[column sep=8em,row sep=5em]
  \ZcC{m} \ar[r,"\pr_m",two heads]\ar[d,dashed, swap, "\tpsi_m"]\ar[dr,"\j_m",near start]
  & \Sym_m(\H)\ar[d,dashed ]\ar[dl,dashed,very near start,"\psi_m"]\\
  H_m(\cms)\ar[r,swap,"\pa{\iota_m}*"hook] & H_m\pa{\cmsD}
 \end{tikzcd}
\end{equation}

@@

Theorem \ref{thm:Hbms*as*ggrep} is then proved in all bidegrees of the form $(m,m)$.

\subsection{General bidegrees $(m-l,m)$.} In this subsection we fix a bidegree $(m-l,m)$
with $0\leq l\leq m$.
Our aim is to prove theorem \ref{thm:Hbms*as*ggrep} for this bidegree.

% \begin{defn}
% %  Let $\tilde\D'$ be the open trapeziod contained in $(0,1)^2\subset\mrS$ which is delimited
% %  by the four points $(1/8;1)$, $(1/4;1/2)$, $(3/4;1/2)$ and $(7/8;1)$ (SEE PICTURE); in particular
% %  $\D\subset\D'$.
% %  Let $\S'\subset\S$ be the closure in $\S$ of $\S\setminus\bar\D$, where $\bar\D$ is the closure of $\D$ in $\S$.
% %  Notice that $\S'$ is also a surface of type $\sg$.
% %  
%  Recall from section \ref{Preliminaries} that $\S'$ is the closure in $\S$ of
%  $\S\setminus\bar\D$, where $\bar\D$ is the closure of $\D$ in $\S$.
%  
%  For all $m$ we define configuration spaces $C_m(\D)$ and $C_m(\S')$ as in definition \ref{defn:cms},
%  using $\D$ and $\mrS'$ instead of $\mrS$, respectively.
%  
%  For every splitting $m=p+(m-p)$ there is an open embedding
%  \[
%  \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
%  \]
%  given by taking the union of configurations:
% 
% %  For the rest of the section let $\gg$ denote the group of isotopy classes of diffeomorphisms
% %  of $\S$ relative to $\partial\S\cup\D'$: in symbols $\gg=\pi_0\pa{\Diff(\S;\partial\S\cup\D'}$.
% \end{defn}
Recall from section \ref{Preliminaries} that $\S'$ is the closure in $\S$ of
$\S\setminus\bar\D$, where $\bar\D$ is the closure of $\D$ in $\S$.

For all $0\leq p\leq m$, the group $\Diff(\S;\partial\S\cup\D)$ acts both on $C_p(\D)\times C_{m-p}(\S')$
and on $\cms$, and the map $\mu$ is equivariant with respect to this action
(see definition \ref{defn:universalSbundle});
hence, using the K\"{u}nneth formula, there is an induced $\gg-$equivariant map in homology
\begin{equation}
 \label{eq:mu*}
 \mu_*\colon H_{p-l}\pa{C_p(\D)}\otimes H_{m-p}\pa{C_{m-p}(\S')}\to H_{m-l}\pa{\cms}.
\end{equation}

@@

We will prove the following lemma, from which theorem \ref{thm:Hbms*as*ggrep} follows:
\begin{lem}
 \label{lem:oplussplitting}
For all $l\leq p\leq m$ the map $\mu_*$ in equation \ref{eq:mu*} is injective, and the collection
of all these maps yields a splitting
\[
 H_{m-l}(\cms)=\bigoplus_{p=l}^m H_{p-l}\pa{C_p(\D)}\otimes H_{m-p}\pa{C_{m-p}(\S')}
\]
\end{lem}
\begin{proof}
Notice that the statement of the lemma doesn't depend on the the action of $\gg$:
we have a map between two vector spaces, we already know that it is $\gg-$equivariant,
we only need to show that it is a bijection. Notice also that lemma \ref{lem:gensymchain}
implies that the two vector spaces have the same dimension.
% 
% We can equivalently consider the maps in cohomology
%  \[
%    \mu^*\colon  H^{m-l}\pa{\cms}\to H^p\pa{C_{m-p}(\S')}\otimes H^{p-l}\pa{C_p(\D)},
%  \]
% and show that they are surjective for $0\leq p\leq m-l$ and together exhibit a product splitting
% \[
%   H^{m-l}(\cms)=\prod_{p=0}^{m-l} H^p\pa{C_{m-p}(\S')}\otimes H^{p-l}\pa{C_p(\D)}.
% \]
% Notice that the natural target for the cohomology map $\mu^*$ is the whole group
% \[
%  H^{m-l}\pa{C_{m-p}(\S')\times C_p(\D)}\simeq \prod_{q=0}^p H^q\pa{C_{m-p}(\S')}\otimes H^{m-l-q}\pa{C_p(\D)},
% \]
% where we have used the K\"unneth formula; we consider here, by abuse of notation,
% the composition of this map with the projection
% on the factor corresponding to $q=p$.
% 
% Using Poincaré-Lefschetz duality and again the K\"unneth formula we reduce to show that the maps
% \[
% \mu^c_* \colon \tH_{m-l}(\cms^c)\to \tH_p\pa{C_{m-p}(\S')^c}\otimes \tH_{p-l}\pa{C_p(\D)^c}
% \]
% are surjective and exhibit $\tH_{m-l}(\cms^c)$ as a product
% \[
% \tH_{m-l}(\cms^c)=\prod_{p=0}^{m-l}\tH_p\pa{C_{m-p}(\S')^c}\otimes \tH_{p-l}\pa{C_p(\D)^c}.
% \]
% Here a few explanations are required:
% we are considering the map $\mu^c\colon \cms^c\to \pa{C_{m-p}(\S')\times C_p(\D)}^c$ which collapses the complement
% of the open subspace $C_{m-p}(\S')\times C_p(\D)\subset\cms$ to the point at infinity; it induces a map
% $\mu^c_*\colon \tH_{m-l}(\cms^c) \to \tH_{m-l}\pa{\pa{C_{m-p}(\S')\times C_p(\D)}^c}$;
% applying Poincaré-Lefschetz duality
% and the K\"unneth formula to the second group we get an isomorphism
% \[
%  \tH_{m-l}\pa{\pa{C_{m-p}(\S')\times C_p(\D)}^c}\simeq \prod_{q=0}^p \tH_q\pa{C_{m-p}(\S')}\otimes\tH_{m-l-q}\pa{C_p(\D)};
% \]
% we are considering the composition of the map $\mu^c_*$ with the projection onto the factor $q=p$, and by
% abuse of notation we call this map simply $\mu^c_*$.
%

Fix $l\leq p\leq m$ and let $[a]=\prod_{j=0}^{\infty}(Q^j\epsilon)^{\alpha_j}$ be a generator of
$H_{p-l}(C_p(\D))$, with $\sum_j\alpha_j=l$ and $\sum_j2^j\alpha_j=p$. Let also
$[b]=\prod_{i=1}^g \u_i^{u_i}\v_i^{v_i}$ be a generator of $H_{m-p}(C_{m-p}(\S'))$, using the
isomorphism proved in the previous subsection, with $\sum_i(u_i+v_i)=m-p$. Here $a$ and
$b$ are chosen singular cycles representing the homology classes, with $a$ supported on $\D$
and $b$ supported on $\S'$.

Then $[a]\otimes [b]$ is a generator of $H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S'))$,
and we are interested in the homology class $\mu_*([a]\otimes [b])$: there is one such
class for any choice of $[a]$ and $[b]$ as above, that is, for any choice of a string of numbers
$(p,(\alpha_j)_{j\geq 0},(u_i,v_i)_{1\leq i\leq g})$
such that $\sum_j\alpha_j=l$, $\sum_j\alpha_j2^j=p$ and $\sum_i(u_i+v_i)=m-p$; our aim
is to show that the collection of all these classes of the form $\mu_*([a]\otimes [b])$
is a basis for $H_{m-l}(\cms)$.

We will study
the intersection of $\mu_*([a]\otimes [b])$ with cohomology classes of $\cms$ represented by
generalised symmetric chains in $\cms^c$.

@@

To compute the algebraic intersection
between $\mu_*([a]\otimes [b])$ and $[p',(\alpha'_j),(u'_i,v'_i)]$ we
consider the map
\[
 \mu^c\colon \cms^c\to \pa{C_p(\D)\times C_{m-p}(\S')}^c
\]
which collapses to $\infty$ the complement in $\cms^c$ of $C_p(\D)\times C_{m-p}(\S')$.

@@

Recall that $\cms^c$ can be filtered according to the norm of cells: a cell $e^{\tup}$ with
$\tup=(l,(x_i),(u_i,v_i))$ has norm $\sum x_i$, and the norm is weakly decreasing along boundaries.
In the previous section we just considered
the associated filtration of the reduced chain complex $\tCh_*(\cms^c)$, whereas now we
consider the closed subcomplex $F_p\cms^c\subset\cms^c$ which is the union
of all cells of norm $\leq p$.

The crucial remark is that $\mu^c$ restricts to a cellular map 
\[
F_p\cms^c\to \pa{C_p(\D)\times C_{m-p}(\S')}^c.
\]
To see this, consider an open cell
cell $e^{\tup}$ for some tuple $\tup=(l, (x_i),(u'_i,v'_i))$ of norm $p'\leq p$:
its intersection with the subspace
$C_p(\D)\times C_{m-p}(\S')$ is empty if $p'<p$, and it is exactly the open product of cells
$e^{\tup'}\times e^{\tup''}$,
where $\tup'=(l,(x_i))$ and $\tup''=(0,(u'_i,v'_i))$, if $p'=p$; therefore
$\mu^c(e^{\tup})$ is $\set{\infty}$ in the first case, and in the second case it
is contained in the union $\set{\infty}\cup e^{\tup'}\times e^{\tup''}$, which is also
contained in the $p'-$skeleton of $\pa{C_p(\D)\times C_{m-p}(\S')}^c$.

% We replace the map $\mu^c$ by a cellular approximation that agrees with it on the $p-$skeleton,
% and by abuse of notation we still call $\mu^c$ the new map.
% 
Consider now any generalised symmetric chain $|p',(\alpha'_j),(u'_i,v'_i)|$ representing
a class in $H_{m+l}(\cms^c)=H^{m-l}(\cms)$, hence $\sum_j\alpha'_j=l$, and suppose $p'\leq p$.

@@

We consider the collection of all strings of the form $(p,(\alpha_j)_{j\geq 0},(u_i,v_i)_{1\leq i\leq g})$
satisfying $\sum\alpha_j=l$,$\sum \alpha_j2^j=p$ and $\sum_i(u_i+v_i)=m-p$;
we order these strings lexicographically,
so that in particular $p$ is weakly increasing; we associate
to each string its corresponding class in $H_{m-l}(\cms)$ of the form $\mu_*([a]\otimes [b])$ and its
corresponding class $[p,(\alpha_j),(u_i,v_i)]\in\tH_{m+l}(\cms^c)$.

Then the matrix of algebraic intersections between these two sets of
classes is an upper-triangular matrix
with $1$'s on the diagonal, and in particular
it is invertible. This shows that the set of classes $\mu_*([a]\otimes [b])$
is a basis for $H_{m-l}(\cms)$.

@@

One could expect that the basis given by all elements $[a]\otimes [b]\in H_{m-l}(\cms)$
is dual to the basis of classes $[p',(\alpha'_j),(u'_i,v'_i)]\in \tH_{m+l}(\cms^c)$, i.e., the
matrix considered in the end of the previous proof is not only upper-triangular but also
diagonal. This is however not true, as the following example shows.

Let $g=1$, $m=2$, $p=1$, $p'=2$ and consider the classes $[a]=\epsilon\in H_1(C_1(\D))$,
$[b]=\u_i\in \H=H_1(C_1(\S'))$; moreover let $[p',(\alpha'_j),u'_1,v'_1]$ be defined by
$p'=2$, $\alpha'_1=1$, $u'_1=v'_1=0$. Represent $[a]$ by a point in $\D$, for example
the point $(1/2,3/4)$; represent $[b]$ by a simple closed curve $b\in\S'$
that intersects only once, transversely, the vertical segment passing through $a$, namely
$\set{1/2}\times(0,1)$.

Then the cycle $|p',(\alpha'_j),u'_1,v'_1|$ consists only of the cell $e^{\tup'}$ with
$\tup'=(1,(x'_1=2),(u'_1=v'_1=0))$, and this cell intersects once, transversely,
the cycle $\mu_*(a\otimes b)$, which is represented by the curve of configurations of two
points in $\mrS$, one of which is fixed at $a$ whereas the other runs along $b$: indeed
there is exactly one position for the second point to lie under $a$.

Hence the algebraic intersection between these two classes is $1$, and since $p'>p$
this is an entry strictly above the diagonal in the matrix considered in the proof
of lemma \ref{lem:oplussplitting}.
